Statistical analysis of periodic data in neuroscience

### 2021-01-01 manuscript/manuscript.Rmd
Minor improvements

nsims <- 10000  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete

@@

The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if not. Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs. However the _multRM_ function in the [_MANOVA.RM_]{https://CRAN.R-project.org/package=MANOVA.RM} package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.

>>>>

nsims <- 100  # global value for the whole script - increase to 100000 for final run, but note this takes around 15 hours to complete

@@

The flowchart in Figure \@ref(fig:flowchart) contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the $T^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, and the $T^2$ statistic if not. Comparisons with more than two conditions should be tested with the $ANOVA^2_{circ}$ statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs. However the _multRM_ function in the [_MANOVA.RM_](https://CRAN.R-project.org/package=MANOVA.RM) package [@Friedrich2019] is able to handle such designs appropriately using a bootstrapping approach.

### 2021-01-01 manuscript/manuscript.tex
Minor improvements

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2020-12-31}

@@

A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's \emph{d} statistic (see e.g. Del Giudice, 2009). This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \emph{HDMD} package \url{https://CRAN.R-project.org/package=HDMD}.

@@

Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic. The project repository contains \emph{R} functions for both of these tests (\emph{anovacirc.test} and \emph{anovacircR.test}).

@@

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs. In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial statistics package \emph{SPSS}.

@@

With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.19). A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001). Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, D = 0.33, \emph{p} \textless{} 0.001). The study by Vilidaite et al. (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders. The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}, and a processed version is included with the \emph{FourierStats} package.

@@

>>>>

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2021-01-01}

@@

A variant of the Mahalanobis distance (the pairwise Mahalanobis distance) can also be used to compute a multivariate measure of effect size, equivalent to Cohen's \emph{d} statistic (see e.g. Del Giudice, 2009). This is a valuable statistic to include when reporting the results of multivariate tests, and an \emph{R} function (\emph{pairwise.mahalanobis}) to calculate it is available as part of the \href{https://CRAN.R-project.org/package=HDMD}{\emph{HDMD}} package.

@@

Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that appropriate multiple comparison correction is applied (e.g.~Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic. The project repository contains an \emph{R} function (\emph{anovacirc.test}) to run both of these tests.

@@

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios using the condition index test. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Many MANOVA implementations cannot deal correctly with random factors (repeated measures) in complex factorial designs. However the \emph{multRM} function in the {[}\emph{MANOVA.RM}{]}\{\url{https://CRAN.R-project.org/package=MANOVA.RM}\} package (Friedrich et al., 2019) is able to handle such designs appropriately using a bootstrapping approach.

@@

With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.19). A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001). Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, D = 0.33, \emph{p} \textless{} 0.001). The study by Vilidaite et al. (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders. The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}, and a processed version is included with the \href{https://github.com/bakerdh/FourierStats}{\emph{FourierStats}} package.

@@

\leavevmode\hypertarget{ref-Friedrich2019}{}%
Friedrich S, Konietschke F, Pauly M. 2019. Resampling-Based Analysis of Multivariate Data and Repeated Measures Designs with the R Package MANOVA.RM. \emph{The R Journal} \textbf{11}:380--400. doi:\href{https://doi.org/10.32614/RJ-2019-051}{10.32614/RJ-2019-051}
